3.3 Cathode Bias Modeling
3.3.1 Physical Background
In a typical self-biased triode stage, a cathode resistor $R_k$ is placed between the cathode and ground to establish the quiescent operating point. To prevent negative feedback from reducing the stage's AC gain, a large bypass capacitor $C_k$ is placed in parallel with $R_k$. However, this capacitor is not a perfect short circuit to all AC signals.

When the tube amplifies a large signal, the asymmetrical nature of the tube's transfer function (or grid conduction on positive peaks) causes the average cathode current to increase. This increased current pumps charge into the bypass capacitor $C_k$. Because the $R_k C_k$ network has a finite time constant (typically 10 to 100 ms), the average cathode voltage $V_k$ slowly rises during a sustained loud note and recovers slowly when the note stops.

Since the grid-to-cathode voltage is $V_{gk} = V_g - V_k$, an increase in $V_k$ makes the grid relatively more negative. This dynamic shift in the operating point has two major sonic consequences:

Dynamic Gain Reduction (Sag): As the operating point shifts into the colder, less conductive region of the triode's characteristic curve, the stage's gain decreases. This introduces a natural, musically pleasing compression known as "sag."
Harmonic Modulation: The shift in bias alters the asymmetry of the transfer function, modifying the harmonic content (particularly enhancing even harmonics) in real-time as the note evolves.
3.3.2 Proposed Mathematical Model
To capture the "sag" and dynamic feel of cathode biasing without resorting to computationally expensive iterative circuit solvers, we propose a phenomenological, feed-forward model implemented in FAUST. Let $x[n]$ denote the normalized input signal.

Step 1 — Half-Wave Rectification: The increase in cathode current occurs predominantly during the positive half-cycles of the input signal. We approximate the charging source using half-wave rectification: $$ x_{rect}[n] = \max(0, x[n]) \tag{6} $$

Step 2 — RC Low-Pass Filtering: The behavior of the $R_k C_k$ network is modeled using a simple first-order IIR low-pass filter, which tracks the envelope of the rectified signal: $$ e_{cath}[n] = e_{cath}[n-1] + \alpha \cdot (x_{rect}[n] - e_{cath}[n-1]) \tag{7} $$ where $\alpha = 1 / \max(1, \tau \cdot f_s \cdot 10^{-3})$, $\tau$ is the RC time constant in milliseconds, and $f_s$ is the sampling rate. This envelope represents the dynamic variation in the cathode voltage $\Delta V_k$.

Step 3 — Operating Point Shift: The envelope is scaled and subtracted from the input of the nonlinear stage. The effective bias shift $\Delta b[n]$ is given by: $$ \Delta b[n] = S_{cath} \cdot e_{cath}[n] \tag{8} $$ where $S_{cath}$ is a scaling factor governing the magnitude of the bias excursion. The finalized triode transfer function becomes: $$ y[n] = \tanh\bigl(a \cdot x[n] + b - \Delta b[n]\bigr) \tag{9} $$ where $a$ is the drive/gain and $b$ is the static bias. Note that subtracting $\Delta b[n]$ shifts the operating point to the left (more negative $V_{gk}$), accurately simulating a colder bias resulting from cathode capacitor charging.

3.3.3 Parameter Analysis and Figure Descriptions
Figure 1 --- Core Functions of the Cathode Bias Model. Figure 1 illustrates the foundational mechanics. Panel (a) shows the half-wave rectification; only the positive peaks (which represent the surges in cathode current) contribute to the charging of the virtual capacitor. Panel (b) visualizes the "Migration Path" on the $\tanh$ transfer function. As the $scale$ factor increases, the operating point (circles) slides along a dotted trajectory into the "Sag" region. Explicit horizontal projections show the resulting drop in normalized output level, providing a clear visual link between the $scale$ parameter and the depth of the dynamic compression (sag) effect.

Figure 2 --- RC Time Constant Dynamics and Musical Feel. Figure 2 captures the temporal "breathing" of the model. Panel (a) shows a rhythmic sequence of guitar-like strikes; the overlaid "Effective Gain" (sag) illustrates how loud picking attacks momentarily "strangle" the stage's headroom, followed by a "note rebound" as the bias recovers. Panel (b) contrasts different circuit topologies: a short time constant ($\tau = 25$ ms) provides a "Tight" feel typical of modern high-gain amps, tracking each strike closely. Conversely, a long time constant ($\tau = 400$ ms, e.g. vintage "Tweed" circuits) creates a "Spongy" feel, where the bias accumulates across the phrase, creating a holistic holistic compression that musician's describe as "breathing."

Figure 3 --- Scale Parameter: Contrast Analysis and Audible Volume Drop. Figure 3 provides an unmistakable visual proof of the $scale$ parameter's influence. Panel (a) compares a symmetric baseline ($scale=0$) with an extreme sag configuration ($scale=4.5$); the resulting "Sag Gap" (red area) shows a flagrant drop in peak-to-peak amplitude. Panel (b) illustrates the resulting progressive compression. By simulating AC-coupling (DC removal), we reveal the true "Audible Volume Drop": as the scale increases, the effective signal energy sags significantly over time (from 0.8 down to 0.35), capturing the physical "choke" effect characteristic of high-gain tube stages.

Figure 4 — Harmonic Spectrum Comparison. The FFT spectrum in Figure 4 compares the steady-state harmonic output of a stage with and without the cathode bias active. The dynamic bias shift fundamentally breaks the odd symmetry of the standard $\tanh$ curve. Consequently, the model with cathode bias engaged exhibits a dramatic increase in even-order harmonics (H2, H4, H6), adding the characteristic musical "warmth" typical of single-ended and non-perfectly-balanced push-pull tube designs.

Figure 5 --- Amplifier Preset Comparison: Dynamic Gain Response. Our model parametrizes physical circuit variations into specific profiles. Figure 5 illustrates how different amplifier archetypes respond to a "Gallop" rhythmic sequence. Panel (a) shows the DC bias trajectories where the memory effect of the RC constant is visible. Panel (b) translates these into the "Dynamic Gain Response" (the musical feel), overlaid with the actual processed waveforms in the background for context. The "Peavey 5150" preset exhibits a tight, rapid recovery typical of modern high-gain designs. Conversely, the "Fender Twin" and "Vox AC30" presets show a slower, more deliberate gain breathing, with the shaded areas highlighting the cumulative volume choke.

Figure 6 --- Multi-Stage Accumulation: The Cascading Choke. A key strength of this methodology is its distributability. Figure 6 simulates a three-stage preamp (configured like a Marshall JCM800), showing the evolution of the "Gain Sag" as the signal propagates through the chain, visualized with background signal textures for each stage. Stage 1 tracks the attack closely, preserving the initial transients ("Tight"). Stage 2, driven harder by the first stage's output, introduces significant dynamic compression ("The Meat"). By Stage 3, the signal is heavily saturated and asymmetrical, resulting in an extreme, long-term "Spongy" recovery. This cascading interaction generates an exceptionally rich, complex, and physically plausible dynamic response that a global static waveshaper cannot replicate.

3.3.4 Model Validity and Limitations
The proposed cathode bias model efficiently captures the fundamental psychoacoustic and dynamic characteristics of physical tube amplifier sag:

Asymmetrical Ck Charging: Approximated successfully via half-wave rectification (Eq. 6).
Time-Variant Gain/Compression: Handled by the first-order RC smoothing filter (Eq. 7).
Dynamic Harmonic Evolution: Captured by shifting the operating point of the nonlinear function (Eqs. 8–9).
By operating strictly in a feed-forward manner, the algorithm avoids zero-delay feedback loops (ZDF) and iterative solvers (e.g., Newton-Raphson), making it extraordinarily cheap to process ($\mathcal{O}(1)$ operations per sample). It compiles efficiently to WebAssembly for real-time, in-browser use.

However, the simplifications introduce minor discrepancies relative to a full SPICE simulation. First, using strict half-wave rectification is an exaggeration; in a physical tube, the cathode current is continuous but highly non-linear. Second, the model ignores the interaction between the anode voltage and the cathode current (the Miller effect and plate resistance variations are not dynamically modeled). Lastly, the decoupling of the grid current model (Section 3.2) and the cathode bias model prevents cross-modulation where severe grid current could further alter the cathode voltage.

Despite these physical simplifications, perceptual validation confirms that the model robustly delivers the coveted "spongy" feel, attack envelope shaping, and harmonic bloom that musicians expect from genuine tube preamplification, all while remaining highly parameterizable.

